New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-25 paper/paper.tex
some improvements

The results show that the execution times are nearly constant for $\symRegRange=52$. Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed, because $\symCount_{53}$ is equal to zero with high probability for realistic cardinalities.

@@

\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated. Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation. $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
\end{itemize}

@@

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation}
\ContinuedFloat
\label{alg:estimate_ml}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symRegRange\gets\dim(\vec{\symCount})-2$
\State $\symRegVal_\text{min} \gets \min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\If{$\symRegVal_\text{min} > \symRegRange$}
\State\Return $\infty$
\EndIf
\State $\symRegVal'_\text{min} \gets \max(\symRegVal_\text{min}, 1)$
\State $\symRegVal_\text{max} \gets \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\State $\symRegVal'_\text{max} \gets \min(\symRegVal_\text{max}, \symRegRange)$
\State $\symZ \gets 0$
\State $\symNumReg'\gets \symCount_{\symRegRange+1}$
\State $\symY \gets 2^{-\symRegVal'_\text{max}}$
\For{$\symRegVal \gets \symRegVal'_\text{max},\symRegVal'_\text{min}$}
\State $\symZ \gets \symZ + \symCount_\symRegVal\cdot\symY$
\Comment here $\symY = 2^{-\symRegVal}$
\State $\symY \gets 2\symY$
\State $\symNumReg'\gets\symNumReg' + \symCount_\symRegVal$
\EndFor
\State $\symNumReg \gets \symNumReg' + \symCount_0$ 
\State $\symCount' \gets \symCount_{\symRegRange+1}$
\If{$\symRegRange\geq 1$}
\State $\symCount' \gets \symCount' + \symCount_{\symRegVal'_\text{max}}$
\EndIf
\State $\symFuncPrime_\text{prev}\gets 0$
\State $\symA\gets \symZ + \symCount_0$
\State $\symB\gets \symZ + 
\symCount_{\symRegRange+1}\cdot 2^{-\symRegRange}$
\If{$\symB \leq 1.5\cdot\symA$}
\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \Comment weak lower bound \eqref{equ:weak_lower_bound}
\Else
\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ \Comment strong lower bound \eqref{equ:strong_lower_bound}
\EndIf
\algstore{myalg}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation (continued)}
\begin{algorithmic}
\algrestore{myalg}
\State $\Delta\symX \gets \symX$
\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$
\State $\symE \gets 1+\lfloor\log_2(\symX)\rfloor$
\State $\symXNormalized \gets
\symX \cdot 2^{-\max(\symRegVal'_\text{max}+1, \symE+2)}$
\Comment $\symXNormalized \in [0, 0.25]$
\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
\State $\symHelper \gets
\symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
\Comment Taylor approximation \eqref{equ:taylor}
\For{$\symRegVal\gets \symE,\symRegVal'_\text{max}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime \gets \symCount'\cdot\symHelper$
\Comment compare \eqref{equ:funcprime}
\For{$\symRegVal\gets(\symRegVal'_\text{max}-1),\symRegVal'_\text{min}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symFuncPrime\gets \symFuncPrime + \symCount_\symRegVal\cdot\symHelper$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime\gets \symFuncPrime + \symX\cdot\symA$
\If{$\symFuncPrime > \symFuncPrime_\text{prev} \wedge \symNumReg' \geq\symFuncPrime$}
\State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
\Comment see \eqref{equ:secant_delta}
\Else
\State $\Delta\symX \gets 0$
\EndIf
\State $\symX \gets \symX + \Delta\symX$
\State $\symFuncPrime_\text{prev}\gets \symFuncPrime$
\EndWhile
\State \Return $\symNumReg\cdot\symX$
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm. \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52}, and \cref{fig:estimation_error_12_14} show very similar results for various HyperLogLog parameters.

What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities. Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:estimation_error_22_10}).

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:estimation_error_22_10}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:estimation_error_12_52}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:estimation_error_12_14}
\end{figure}

\subsection{Performance}
We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. The average computation time shows a different behavior over the cardinality than those of the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}). As can be seen, the average calculation time is longer for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}

>>>>

The results show that the execution times are nearly constant for $\symRegRange=52$. Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers ($\symCount_{53}>0$) for realistic cardinalities.

@@

\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated. Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor series approximation. $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
\end{itemize}

@@

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation.}
\ContinuedFloat
\label{alg:estimate_ml}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symRegRange\gets\dim(\vec{\symCount})-2$
\State $\symRegVal_\text{min} \gets \min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\If{$\symRegVal_\text{min} > \symRegRange$}
\State\Return $\infty$
\EndIf
\State $\symRegVal'_\text{min} \gets \max(\symRegVal_\text{min}, 1)$
\State $\symRegVal_\text{max} \gets \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\State $\symRegVal'_\text{max} \gets \min(\symRegVal_\text{max}, \symRegRange)$
\State $\symZ \gets 0$
\State $\symNumReg'\gets \symCount_{\symRegRange+1}$
\State $\symY \gets 2^{-\symRegVal'_\text{max}}$
\For{$\symRegVal \gets \symRegVal'_\text{max},\symRegVal'_\text{min}$}
\State $\symZ \gets \symZ + \symCount_\symRegVal\cdot\symY$
\Comment here $\symY = 2^{-\symRegVal}$
\State $\symY \gets 2\symY$
\State $\symNumReg'\gets\symNumReg' + \symCount_\symRegVal$
\EndFor
\State $\symNumReg \gets \symNumReg' + \symCount_0$ 
\State $\symCount' \gets \symCount_{\symRegRange+1}$
\If{$\symRegRange\geq 1$}
\State $\symCount' \gets \symCount' + \symCount_{\symRegVal'_\text{max}}$
\EndIf
\State $\symFuncPrime_\text{prev}\gets 0$
\State $\symA\gets \symZ + \symCount_0$
\State $\symB\gets \symZ + 
\symCount_{\symRegRange+1}\cdot 2^{-\symRegRange}$
\If{$\symB \leq 1.5\cdot\symA$}
\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \Comment weak lower bound \eqref{equ:weak_lower_bound}
\Else
\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ \Comment strong lower bound \eqref{equ:strong_lower_bound}
\EndIf
\algstore{myalg}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation (continued).}
\begin{algorithmic}
\algrestore{myalg}
\State $\Delta\symX \gets \symX$
\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$
\State $\symE \gets 1+\lfloor\log_2(\symX)\rfloor$
\State $\symXNormalized \gets
\symX \cdot 2^{-\max(\symRegVal'_\text{max}+1, \symE+2)}$
\Comment $\symXNormalized \in [0, 0.25]$
\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
\State $\symHelper \gets
\symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
\Comment Taylor approximation \eqref{equ:taylor}
\For{$\symRegVal\gets \symE,\symRegVal'_\text{max}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime \gets \symCount'\cdot\symHelper$
\Comment compare \eqref{equ:funcprime}
\For{$\symRegVal\gets(\symRegVal'_\text{max}-1),\symRegVal'_\text{min}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symFuncPrime\gets \symFuncPrime + \symCount_\symRegVal\cdot\symHelper$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime\gets \symFuncPrime + \symX\cdot\symA$
\If{$\symFuncPrime > \symFuncPrime_\text{prev} \wedge \symNumReg' \geq\symFuncPrime$}
\State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
\Comment see \eqref{equ:secant_delta}
\Else
\State $\Delta\symX \gets 0$
\EndIf
\State $\symX \gets \symX + \Delta\symX$
\State $\symFuncPrime_\text{prev}\gets \symFuncPrime$
\EndWhile
\State \Return $\symNumReg\cdot\symX$
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm. \cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24}, \cref{fig:max_likelihood_estimation_error_16_16},\cref{fig:max_likelihood_estimation_error_22_10},\cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.

What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities. Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:max_likelihood_estimation_error_12_20}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:max_likelihood_estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:max_likelihood_estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:max_likelihood_estimation_error_22_10}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:max_likelihood_estimation_error_12_52}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:max_likelihood_estimation_error_12_14}
\end{figure}

\subsection{Performance}
We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}). As can be seen, the average execution time is larger for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}
